Spatiotemporal Optimization of Cosmological Particel-Mesh Simulation

### 2022-12-02 docs/papers/sto/sto.tex
FIXUP draft: add so methods

%FIXME copied from ../adjoint/adjoint.tex
\newcommand{\deltaD}{\delta^\textsc{d}}
\newcommand{\deltaK}{\delta^\textsc{k}}
\renewcommand{\d}{d}
\newcommand{\p}{\partial}
\newcommand{\cJ}{\mathcal{J}}
\newcommand{\cR}{\mathcal{R}}
\newcommand{\cL}{\mathcal{L}}
\newcommand{\cH}{\mathcal{H}}
\bmdefine{\vzero}{0}
\bmdefine{\vI}{I}
\bmdefine{\vnabla}{\nabla}
\bmdefine{\vtheta}{\theta}  % parameters
\bmdefine{\vomega}{\omega}  % white noise modes
\bmdefine{\vk}{k}  % wavevectors
\bmdefine{\vx}{x}  % comoving and canonical coordinates
\bmdefine{\vq}{q}  % Lagrangian coordinates
\bmdefine{\vs}{s}  % displacements
\bmdefine{\vp}{p}  % canonical momenta
\bmdefine{\vv}{v}  % FIXME some velocities
\bmdefine{\va}{a}  % accelerations
\bmdefine{\vz}{z}  % states
\bmdefine{\vf}{f}
\bmdefine{\vF}{F}
\bmdefine{\vDelta}{\Delta}
\bmdefine{\vLambda}{\Lambda}
\bmdefine{\vlambda}{\lambda}
\bmdefine{\vvarphi}{\varphi}
\bmdefine{\vxi}{\xi}  % x adjoint
\bmdefine{\vpi}{\pi}  % p adjoint
\bmdefine{\valpha}{\alpha}  % force vjp x gradient
\bmdefine{\vzeta}{\zeta}  % force vjp theta gradient
\newcommand{\half}{\nicefrac12}
\newcommand{\As}{A_\mathrm{s}}
\newcommand{\ns}{n_\mathrm{s}}
\newcommand{\Omegam}{\Omega_\mathrm{m}}
\newcommand{\Omegab}{\Omega_\mathrm{b}}
\newcommand{\Mpc}{\mathrm{Mpc}}
\newcommand{\ic}{\mathrm{i}}
\newcommand{\Plin}{P_\mathrm{lin}}

@@

Again combine Lagrangian and Eulernian fields.

@@

@@

@@

\begin{table*}
  \centering
  \caption{Ranges of GADGET-4 and \pmwd\ configuration and cosmological
    parameters.
    Note that the grid ratio need next fast len to determine the mesh
    shape for fast FFT.
    Given the box size, the mesh shape determines the cell size.
    The softening parameter gives the ratio of the comving softening
    length to the mean particle spacing.
    The curvature $\Omega_k$ is related to the separate universe
    simulation.
    We sample parameters applicable to GADGET-4 during data generation,
    and those applicable to \pmwd\ during training.
    \textsuperscript\dag The box size is determined jointly by the
    \pmwd\ mesh shape and mesh cell size below, which are assumed to be
    sampled independently.
    In practice, we sample one conditioned on the other and the box
    size.
    * Number of time steps from beginning to end ...
  }
  \label{tab:param}
  \begin{tabular}{lcccr}
    \toprule
    applicability & parameter & distribution & range & sampling \\
    \midrule
    & $128^3$ white noise & $\mathcal{N}(\vzero, \vI)$ & 512 realizations & MC \\
    \cmidrule(lr){2-5}
    & 121 snapshots, $a$ & uniform & [1/16, 1] & deterministic \\
    \cmidrule(lr){2-5}
    & box size in Mpc\textsuperscript\dag & log-trapezoidal & [25.6, 2560) \\
    \cmidrule(lr){2-4}
    & snapshot offset, $\Delta\!a$ & uniform & [0, 1/128) \\
    \cmidrule(lr){2-4}
    GADGET-4 & $A_\mathrm{s} \times 10^9$ & log-uniform & [1, 4) \\
    \cmidrule(lr){2-4}
    \& \pmwd\ & $n_\mathrm{s}$ & log-uniform & [0.75, 1.25) & RQMC \\
    \cmidrule(lr){2-4}
    & $\Omega_\mathrm{m}$ & log-uniform & [1/5, 1/2) & see \autoref{fig:sobol} \\
    \cmidrule(lr){2-4}
    & $\Omega_\mathrm{b} / \Omega_\mathrm{m}$ & log-uniform & [1/16, 1/2) \\
    \cmidrule(lr){2-4}
    & $\Omega_k / (1 - \Omega_k)$ & uniform & [-0.5, 0.5) \\
    \cmidrule(lr){2-4}
    & $h$ & log-uniform & [0.5, 1) \\
    \cmidrule(lr){1-4}
    GADGET-4 & softening ratio & log-uniform & [1/50, 1/20) \\
    \cmidrule(lr){1-5}
    & mesh shape & log-uniform & [128, 512] \\
    \cmidrule(lr){2-4}
    \pmwd\ & mesh cell size in Mpc & log-uniform & [0.2, 5] & MC \\
    \cmidrule(lr){2-4}
    & number of time steps* & log-uniform & [10, 1000] \\
    \bottomrule
  \end{tabular}
\end{table*}

>>>>

%FIXME copied from ../adjoint/adjoint.tex
\newcommand{\deltaD}{\delta^\textsc{d}}
\newcommand{\deltaK}{\delta^\textsc{k}}
\renewcommand{\d}{d}
\newcommand{\p}{\partial}
\newcommand{\cJ}{\mathcal{J}}
\newcommand{\cR}{\mathcal{R}}
\newcommand{\cL}{\mathcal{L}}
\newcommand{\cH}{\mathcal{H}}
\bmdefine{\vzero}{0}
\bmdefine{\vI}{I}
\bmdefine{\vnabla}{\nabla}
\bmdefine{\vtheta}{\theta}  % parameters
\bmdefine{\vvartheta}{\vartheta}  % NN parameters
\bmdefine{\vomega}{\omega}  % white noise modes
\bmdefine{\vk}{k}  % wavevectors
\bmdefine{\vx}{x}  % comoving and canonical coordinates
\bmdefine{\vq}{q}  % Lagrangian coordinates
\bmdefine{\vs}{s}  % displacements
\bmdefine{\vp}{p}  % canonical momenta
\bmdefine{\vv}{v}  % FIXME some velocities
\bmdefine{\va}{a}  % accelerations
\bmdefine{\vz}{z}  % states
\bmdefine{\vf}{f}
\bmdefine{\vF}{F}
\bmdefine{\vDelta}{\Delta}
\bmdefine{\vLambda}{\Lambda}
\bmdefine{\vlambda}{\lambda}
\bmdefine{\vvarphi}{\varphi}
\bmdefine{\vxi}{\xi}  % x adjoint
\bmdefine{\vpi}{\pi}  % p adjoint
\bmdefine{\valpha}{\alpha}  % force vjp x gradient
\bmdefine{\vzeta}{\zeta}  % force vjp theta gradient
\newcommand{\half}{\nicefrac12}
\newcommand{\As}{A_\mathrm{s}}
\newcommand{\ns}{n_\mathrm{s}}
\newcommand{\Omegam}{\Omega_\mathrm{m}}
\newcommand{\Omegab}{\Omega_\mathrm{b}}
\newcommand{\Mpc}{\mathrm{Mpc}}
\newcommand{\ic}{\mathrm{i}}
\newcommand{\linear}{\mathrm{lin}}
\newcommand{\tophat}{\mathrm{TH}}
\newcommand{\gauss}{\mathrm{G}}

@@

Again combine Lagrangian and Eulerian fields.

@@

An alternative loss function can be the Eulernian density weighted
particle displacement MSE.

@@

The dimensionless power spectrum
%
\begin{equation}
\Delta^2(k) \triangleq \frac{k^3 P(k)}{2 \pi^2}.
\end{equation}

Variance of linear matter overdensity in a top-hat window
%
\begin{equation}
\sigma_\tophat^2(R) = \int_0^\infty \frac{\d k}k
  \Delta_\linear^2(k) W_\tophat^2(kR),
\end{equation}
where $W_\tophat(kR) = \cdots$ is the Fourier transform of the top-hat
window.
Likewise $\sigma_\gauss^2(R)$ is defined with a Gaussian window
$W_\gauss(kR) = \cdots$.

Various nonlinear scales are
\begin{itemize}
\item $k_P^{-1}$: determined by $\Delta_\linear^2(k_P) \triangleq 1$;
\item $R_\tophat$: determined by $\sigma_\tophat^2(R_\tophat) \triangleq
  1$;
\item $R_\gauss$: determined by $\sigma_\gauss^2(R_\tophat) \triangleq
  1$;
\item $R_\mathrm{d}$: rms linear theory displacement;
\item $\d R / \d\ln a$: time derivatives of the above scales.
\end{itemize}

Other relevant scales are
\begin{itemize}
\item $l_\mathrm{cell}$: PM cell size.
\item particle spacing;
\item softening length.
\end{itemize}

Relevant scale-independent dimensionless factors are
\begin{itemize}
\item $D(a)$: linear growth factor;
\item $f(a)$: linear growth rate $f \triangleq \d\ln D / \d\ln a$;
\item $\Omegam(a)$: matter density parameter at $a$;
\item $\d\ln H / \d\ln a$: time derivative of Hubble expansion;
\item $a$: scale factor;
\item $\Delta a$: time step size in $a$.
\end{itemize}

We combine the above list of parameters in $\vvartheta$.

Force sharpening
%
\begin{equation}
\frac{k_i}{k^2} \to \frac{k_i}{k^2}
  h(k_i; \vvartheta) g(k; \vvartheta) \prod_{j=1}^3 f(k_j; \vvartheta),
\end{equation}
%
where $f$, $g$, $h$ are neural networks, each approximating a nonlinear
function of the dimensionless features, e.g.,
%
\begin{equation}
g(k; \vvartheta) = g(k/k_P, kR_\tophat, \cdots, D, f, \cdots).
\end{equation}

@@

\begin{table*}
\centering
\caption{Ranges of GADGET-4 and \pmwd\ configuration and cosmological
parameters.
Note that the grid ratio need next fast len to determine the mesh shape
for fast FFT.
Given the box size, the mesh shape determines the cell size.
The softening parameter gives the ratio of the comving softening length
to the mean particle spacing.
The curvature $\Omega_k$ is related to the separate universe simulation.
We sample parameters applicable to GADGET-4 during data generation, and
those applicable to \pmwd\ during training.
\textsuperscript\dag The box size is determined jointly by the \pmwd\
mesh shape and mesh cell size below, which are assumed to be sampled
independently.
In practice, we sample one conditioned on the other and the box size.
* Number of time steps from $a=1/16$ to $a=1$.
}
\label{tab:param}
\begin{tabular}{lcccr}
  \toprule
  applicability & parameter & distribution & range & sampling \\
  \midrule
  & $128^3$ white noise & $\mathcal{N}(\vzero, \vI)$ & 512 realizations & MC \\
  \cmidrule(lr){2-5}
  & 121 snapshots, $a$ & uniform & [1/16, 1] & deterministic \\
  \cmidrule(lr){2-5}
  & box size in Mpc\textsuperscript\dag & log-trapezoidal & [25.6, 2560) \\
  \cmidrule(lr){2-4}
  & snapshot offset, $\Delta\!a$ & uniform & [0, 1/128) \\
  \cmidrule(lr){2-4}
  GADGET-4 & $\As \times 10^9$ & log-uniform & [1, 4) \\
  \cmidrule(lr){2-4}
  \& \pmwd\ & $\ns$ & log-uniform & [0.75, 1.25) & RQMC \\
  \cmidrule(lr){2-4}
  & $\Omegam$ & log-uniform & [1/5, 1/2) & see \autoref{fig:sobol} \\
  \cmidrule(lr){2-4}
  & $\Omegab / \Omegam$ & log-uniform & [1/16, 1/2) \\
  \cmidrule(lr){2-4}
  & $\Omega_k / (1 - \Omega_k)$ & uniform & [-0.5, 0.5) \\
  \cmidrule(lr){2-4}
  & $h$ & log-uniform & [0.5, 1) \\
  \cmidrule(lr){1-4}
  GADGET-4 & softening ratio & log-uniform & [1/50, 1/20) \\
  \cmidrule(lr){1-5}
  & mesh shape & log-uniform & [128, 512] \\
  \cmidrule(lr){2-4}
  \pmwd\ & mesh cell size in Mpc & log-uniform & [0.2, 5] & MC \\
  \cmidrule(lr){2-4}
  & number of time steps* & log-uniform & [10, 1000] \\
  \bottomrule
\end{tabular}
\end{table*}
